If the Null Fits, You Must Omit: Measurement Artifacts Undermine Group Polarization Replications

### 2025-10-26 GroupPolarizationStatmod/paper/intro-latest.tex
finished intro draft

If an extremist’s opinion falls and it is measured with a Likert scale, will
it make a noise? The answer is often no~\cite{Liddell2018,Turner2020}. This
means a group’s opinions can appear to radicalize over time even when no real
shift occurs. The problem matters because Likert and other ordinal scales are
ubiquitous in research on opinion change and radicalization across the social
and cognitive sciences~\cite{Liddell2018}. One influential framework in this
area is group polarization—the idea that discussion among like-minded people
pushes opinions toward extremes, producing ``echo chambers.'' Cass Sunstein, a
Harvard Law professor and former White House administrator, has written
extensively on the topic~\cite{Sunstein2019}, claiming that group
polarization is a scientific law—a regular, quantitative relationship that
will always be observed when certain assumptions are
met~\cite{Cartwright1999}. Yet Roger Brown himself noted in Social Psychology
that group polarization ``does not occur with every group'' and that ``the
effect is not large.'' Given the influence of this theory and the hundreds of
claimed ``replications,'' it is essential to ask whether many of these findings
may instead be measurement artifacts—spurious effects created when the
``neutral'' point of the measurement scale fails to align with the group’s own
center of opinion.

@@

We selected ten representative journal articles with sixty total experimental
conditions among them. For a representative sample of group polarization
research across decades, we selected two articles for each of five article
categories: \emph{Social Comparisons}, \emph{Persuasive Arguments},
\emph{Self-Categorization}, \emph{Social Decisions}, and \emph{Politics}. The
first four of these categories reference the theoretical tradition that the
authors were trying to prove right—the fifth, \emph{Politics}, was the
subject matter of the studies, but the two studies took no clear side in the
group polarization theory wars. This resulted in a representative sample of
group polarization research over time, as well: one article is from 1969,
four are from the 1970s, . Information about our corpus is summarized below
(Table~\ref{tab:articles}).  Please see the Supplemental Information

% \begin{table}[ht]

%   \caption{\textbf{Journal articles in our corpus.}}
%   \label{tab:articles}

%   \centering 
%   \rowcolors{2}{white}{gray!7}
%   \begin{tabular}{m{1.35in} m{2.45in}
%     m{1.3in} m{0.85in}}

%     \toprule

%     \textbf{Authors (Year)}& \textbf{Title} & \textbf{Journal} &
%     \textbf{Category} \\

%     \midrule 
%     Moscovici \& Zavalloni (1969) & The Group as a Polarizer of
%     Attitudes & \emph{JPSP} & Politics \\ 
%     Schkade, Sunstein, \& Hastie (2010) & When
%     Deliberation Produces Extremism & \emph{Critical~Review} & Politics \\

%     Bishop \& Myers (1970) & Discussion Effects on Racial Attitudes &
%     \emph{Science} & Social Comp's \\ 
%     Myers (1975) & Discussion-Induced Attitude Polarization & 
%     \emph{Human Relations} & Social Comp's \\

%     Burnstein \& Vinokur (1973) & Testing Two Classes of Theories About Group
%     Induced Shifts In Individual Choice & \emph{JESP} & Persuasive Arg's \\
%     Burnstein \& Vinokur (1975) & What a person thinks upon learning he has
%     chosen differently from others: Nice evidence for the
%     persuasive-arguments\ldots & \emph{JESP} & Persuasive Arg's \\

%     Abrams, et al., (1990) & Knowing What to Think by Knowing Who You Are:
%     Self‐Categorization and the Nature of Norm Formation\ldots & 
%     \emph{Brit. J. of Soc. Psych.} & Self-Categ'n \\
%     Hogg, Turner, \& Davidson (1990) & Polarized Norms and Social Frames of 
%     Reference: A Test of the Self-Categorization Theory of Group 
%     Polarization & \emph{B. \& Appl. Soc. Psych.} & Self-Categ'n \\
%     Krizan \& Baron (2007) & Group Polarization and Choice Dilemmas: How
%     Important is Self-Categorization? & \emph{Eur. J. of Soc. Psych.} &
%     Self-Categ'n \\

%     Friedkin (1999) & Choice Shifts \& Group Polarization & 
%     \emph{Amer. Soc.  Rev.} & Soc'l Decisions \\

% \bottomrule
% \end{tabular}
% \end{table}
\begin{table}[ht]
\centering
\caption{\textbf{Journal articles in our corpus.}}
\label{tab:articles}

\begin{tabular}{m{0.9in} m{1.35in} m{2.45in} m{1.3in}}
\toprule
\textbf{Category} & \textbf{Authors (Year)} & \textbf{Title} & \textbf{Journal} \\
\midrule

\rowcolor{gray!10}
\multirowcell{2}{Politics} 
  & Moscovici \& Zavalloni (1969) & The Group as a Polarizer of Attitudes & \emph{JPSP} \\
\rowcolor{gray!10}
  & Schkade, Sunstein, \& Hastie (2010) & When Deliberation Produces Extremism & \emph{Critical~Review} \\[0.2em]

\rowcolor{gray!5}
\multirowcell{2}{Social Comp's \vspace{0.7em}}
% \multirowcell{2}{Social Comp's}
  & Bishop \& Myers (1970) & Discussion Effects on Racial Attitudes & \emph{Science} \\[0.2em]
\rowcolor{gray!5}
  & Myers (1975) & Discussion-Induced Attitude Polarization & \emph{Human Relations} \\[0.3em]

\rowcolor{gray!10}
\multirowcell{2}{Persuasive Arg's}
  & Burnstein \& Vinokur (1973) & Testing Two Classes of Theories About Group Induced Shifts In Individual Choice & \emph{JESP} \\
\rowcolor{gray!10}
  & Burnstein \& Vinokur (1975) & What a person thinks upon learning he has chosen differently from others: Nice evidence for the persuasive-arguments\ldots & \emph{JESP} \\

\rowcolor{gray!5}
\multirowcell{3}{Self-Categ'n}
  & Abrams, et al., (1990) & Knowing What to Think by Knowing Who You Are: Self‐Categorization and the Nature of Norm Formation\ldots & \emph{Brit. J. of Soc. Psy.} \\
\rowcolor{gray!5}
  & Hogg, Turner, \& Davidson (1990) & Polarized Norms and Social Frames of Reference: A Test of the Self-Categorization Theory of Group Polarization & \emph{B. \& Appl. Soc. Psy.} \\
\rowcolor{gray!5}
  & Krizan \& Baron (2007) & Group Polarization and Choice Dilemmas: How Important is Self-Categorization? & \emph{Eur. J. of Soc. Psy.} \\

\rowcolor{gray!10}
\multirowcell{1}{Soc'l Decisions}
  & Friedkin (1999) & Choice Shifts \& Group Polarization & \emph{Amer. Soc. Rev.} \\

\bottomrule
\end{tabular}
\end{table}

\subsection{Measurement procedure}

@@

Note how ordinal scales clip all latent opinions with a magnitude greater than the
magnitude of its extreme values. 
lead to false positives because ordinal scales cannot track changes in
extreme opinions beyond the ceiling of the measurement scale—this approach is
ubiquitous in social psychology and survey research, not at all limited to
group polarization.

>>>>

If an extremist’s opinion falls and it is measured with a Likert scale, will
it make a noise? The answer is often no~\cite{Liddell2018,Turner2020}. This
means a group’s opinions can appear to radicalize over time even when no real
shift occurs. The problem matters because Likert and other ordinal scales are
ubiquitous in research on opinion change and radicalization across the social
and cognitive sciences~\cite{Liddell2018}. One influential framework in this
area is group polarization—the idea that discussion among like-minded people
pushes opinions toward extremes, producing ``echo chambers.'' Cass Sunstein, a
Harvard Law professor and former White House administrator, has written
extensively on the topic~\cite{Sunstein2019}, claiming that group
polarization is a scientific law—a regular, quantitative relationship that
will always be observed when certain assumptions are
met~\cite{Cartwright1999}. Yet Roger Brown himself noted in Social Psychology
that group polarization ``does not occur with every group'' and that ``the
effect is not large''~\cite{Brown1986}. 
Given the influence of this theory and the hundreds of claimed ``replications'' of
group polarization, it is essential to ask whether any of these replications may
instead be \emph{measurement artifacts}—spurious effects created when the
``neutral'' point of the measurement scale fails to align with the group’s own
center of opinion. The science of opinion change has a central role in designing new
methods for social organization and social change to face pressing existential
challenges~\cite{Galesic2021,Galesic2023}.

@@

We selected ten representative journal articles with sixty total experimental
conditions among them. For a representative sample of group polarization
research across decades, we two articles for three of the five article
categories: \emph{Social Comparisons}, \emph{Persuasive Arguments},
 and \emph{Politics}. We included three \emph{Self-Categorization} studies and one \emph{Social Decisions}.
All but the \emph{Politics} category refers to the theoretical tradition that the
authors were trying to prove right—the two \emph{Politics} articles were agnostic
to group psychology theories. This resulted in a representative sample research
over time, as well: we have one study from 1969,
four from the 1970s, three from the 1990s, one from 2007 and one from 2010. 
Information about our corpus is summarized below with references 
(Table~\ref{tab:articles}).

\include{article-table.tex}

Each study used subtly different experimental methods that make it difficult
to build cumulative general knowledge. But all of the studies share one thing
in common: they all measure participant opinions on an ordinal scale. We start
with a non-technical review of how ordinal opinion measurements work in group
polarization, and then how the distributions of measurements are analyzed to
determine if group opinions became polarized. Then we can explain our
close-reading analysis to extract the necessary experimental design parameters
we need to simulate all the experiments in our corpus.

The corpus contains
sixty different experimental conditions in the ten corpus journal articles. We 
simulate experiments for all 57 experimental conditions reported to have 
replicated group polarization, out of 60 total conditions (three were
null control conditions).

Using simulations we will estimate the number of these 57 replications that
must be retracted because null-polarization simulations appear polarized due
to measurement artifacts. We collect experiment design parameters that fill out 
details of the simulation model of group polarization experiments.

\subsection{Ordinal Opinion Measurements: Problems and Simulations}

@@

Note that ordinal scales clip all latent opinions with a magnitude greater
than the magnitude of its extreme values. This can make null-polarization
outcomes look like group polarization. Consider the following example 
with a group of just two people to see why: say they start with latent
opinions of 2.1 and 4.7, which they report as +2 and +3, let's say, on Moscovici
and Zavalloni's scale that tops out at +3. If they then discuss the topic and their
opinions move toward one another by 1 each, then their latent opinions would be 3.1
and 3.7. Before and after discussion the latent mean was a constant 3.4. 
When reported on an ordinal scale, both opinions are now +3—the average measured
opinion \emph{spuriously increased} from 2.75 to 3.0!

To determine whether a replication must be retracted, we developed a model to
simulate measurements of null-polarizations, labelled \emph{mere agreement}, where
the mean group opinion stays constant, but variance decreases. In other words,
simulated participants report more agreement following discussion, 
but they do not become radicalized.

The simulations are seeded by actual group polarization experiment parameters.
We copied the pre- and post-discussion opinion means reported in each article.
There is no original data available for any of the experiments. Sometimes more
information was available about the pre- and post-discussion
distributions—we used this information to decide whether replications plausibly 
have null-polarization. We recorded all available data with a guide to finding it in
its original article in the file `StudiesAnalysis.csv`, available with the
Supplemental Information or to view/download on
GitHub\footnote{\url{https://github.com/mt-digital/gp-statmod/GroupPolarizationStatmod/data/StudiesAnalysis.csv}}.

\subsection{Research Overview}

We collected original data across 10 journal articles to show that 95\% of 
the replications reported must be retracted due to non-identifiability with a
null-polarization model. These replications cannot be salvaged because the
original data is no longer available, and thus cannot be re-analyzed with a 
better statistical model that properly accounts for the measurement procedure.

In the Analysis we go further: we use simulations to show that 
the experimental designs in our corpus are beset by high false positive rates 
\emph{even with a better statistical model}. The only way to reduce false
positive rates with the experimental designs used in the papers is to increase
the significance threshold (in terms of Cohen's $d$ as explained in more
detail in the Methods).

Group polarization was taken as one of the most reliable group effects in social
science—but we showed it is impossible to distinguish 95\% of group
polarization detections from a null-polarization model. This means 54
replications must be retracted—and recall that this was a selection of the
highest quality, most representative group polarization research. 
To understand group polarization, new
methods will be required.

This sort of error threatens our ability to construct a
cumulative science of opinion change, social influence, and social change
generally. We need reliable social science as the foundation for new
strategies for social organization capable of reversing decline and powering
sustainable futures. We have a long way to go.
